---
title: "Generating synthetic scRNA-seq samples with scStable"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating synthetic scRNA-seq samples with scStable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

**scStable** generates multiple synthetic scRNA-seq samples from a single
scRNA-seq sample (the *scRNA-seq reference*). All synthetic samples keep the
same cells as the reference and differ in gene expression by between-sample
variation learned from multi-sample bulk RNA-seq data.

* **Reference mode**: a tissue- and condition-matched bulk RNA-seq reference is
  available.
* **Bulk-reference-free mode**: pre-estimated tissue-specific parameters (for
  example from GTEx) are used instead.

```{r setup, eval = FALSE}
library(scStable)
bulkRNA_matrix <- readRDS(system.file("extdata", "bulk_example.rds", package = "scStable"))
scRNA_matrix   <- readRDS(system.file("extdata", "sc_example.rds",   package = "scStable"))
```

## Preprocessing

Select genes shared by both references, highly variable genes, and the top
k = 10 PCs of the scRNA-seq reference.

```{r prep, eval = FALSE}
prep <- synthreplicate_prep(bulkRNA_matrix, scRNA_matrix, number.gene = 1500)
```

## Step 1: model the scRNA-seq reference

This fits a cell-label-free scDesign3 model (NB marginals, Gaussian copula,
PCs as covariates).

```{r scfit, eval = FALSE}
sc_fit <- scDesign3_fit(prep$sc, prep$pca, save_dir = "scStable_model")
```

## Step 2: model the bulk RNA-seq reference

```{r fitbulk, eval = FALSE}
bulk_fit <- fit_bulk(prep$bulk, n_cores = 4)
```

## Step 3: map between-sample variation

Draw R synthetic bulk samples. The default `use.cor = NULL` uses the full
covariance for 30 or more bulk samples and a diagonal covariance otherwise.

```{r genbulk, eval = FALSE}
bulk_synth <- synthreplicate_gen_bulk(prep$bulk, bulk_fit$mu, bulk_fit$cov,
                                      bulk_fit$optimal_c, prep$sc,
                                      number.replicate = 100)
```

## Step 4: generate synthetic scRNA-seq samples

`scaling_factor` is the scale factor λ: `0` injects no variation, `1` injects
realistic variation, and values above `1` amplify it.

```{r gensc, eval = FALSE}
files <- synthreplicate_gen_sc(prep$bulk, bulk_synth$sampling_bulk, bulk_fit$mu,
                               bulk_synth$d, bulk_fit$optimal_c, prep$sc, sc_fit,
                               scaling_factor = 1, save.dir = "scStable_samples")
```

## Bulk-reference-free mode

Save the parameters once per tissue, then reuse them:

```{r free, eval = FALSE}
saveRDS(fit_bulk(bulkRNA_matrix, n_cores = 4), "blood_params.rds")

res <- synthreplicate_from_tissue(
  scRNA_matrix     = scRNA_matrix,
  bulk_params      = "blood_params.rds",
  save_dir         = "scStable_model",
  replicate_dir    = "scStable_samples_free",
  number.replicate = 100
)
```
