---
title: "Generating synthetic scRNA-seq replicates with scStable"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating synthetic scRNA-seq replicates with scStable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

**scStable** generates synthetic single-cell RNA-seq (scRNA-seq) replicates that
are anchored to bulk RNA-seq data. The method (1) fits a multivariate model to a
reference bulk RNA-seq matrix, (2) samples synthetic bulk profiles, and (3)
propagates the induced per-gene variation into newly simulated single-cell count
matrices using `scDesign3`.

The pipeline has five steps, each with a dedicated function.

```{r setup}
library(scStable)
```

## Inputs

You need two matrices with **gene symbols as `rownames`**:

* `bulkRNA_matrix` — bulk RNA-seq counts, genes x samples (normalized).
* `scRNA_matrix` — reference single-cell counts, genes x cells.

## Step 1 — Preprocess

```{r prep}
prep <- synthreplicate_prep(
  bulkRNA_matrix = bulkRNA_matrix,
  scRNA_matrix   = scRNA_matrix,
  number.pc      = 10,
  number.gene    = 1500
)
```

## Step 2 — Fit the bulk distribution

```{r fitbulk}
bulk_fit <- fit_bulk(
  bulkRNA_matrix = prep$bulk,
  n_cores        = 20
)
```

## Step 3 — Fit the single-cell model

```{r scfit}
sc_fit <- scDesign3_fit(
  scRNA_matrix = prep$sc,
  top_pcs      = prep$pca,
  save_dir     = "scStable_model",
  use.option   = 2
)
```

## Step 4 — Sample synthetic bulk replicates

```{r genbulk}
bulk_synth <- synthreplicate_gen_bulk(
  bulkRNA_matrix   = prep$bulk,
  mu               = bulk_fit$mu,
  cov              = bulk_fit$cov,
  optimal_c        = bulk_fit$optimal_c,
  scRNA_matrix     = prep$sc,
  use.cor          = 3,
  number.replicate = 100
)
```

## Step 5 — Generate synthetic single-cell replicates

```{r gensc}
synthreplicate_gen_sc(
  bulkRNA_matrix = prep$bulk,
  bulk_synth     = bulk_synth$sampling_bulk,
  mu             = bulk_fit$mu,
  d              = bulk_synth$d,
  optimal_c      = bulk_fit$optimal_c,
  scRNA_matrix   = prep$sc,
  scRNA_list     = sc_fit,
  match.option   = 2,
  save.dir       = "scStable_replicates"
)
```

Synthetic replicates are written to `scStable_replicates/replicate<i>.csv`.

## One-call wrapper

```{r wrapper}
res <- synthreplicate_from_tissue(
  tissue_name      = "Liver",
  scRNA_matrix     = scRNA_matrix,
  gtex_data_dir    = "path/to/GTEx/tissue_data/",
  save_dir         = "scStable_model",
  replicate_dir    = "scStable_replicates",
  number.replicate = 100
)
```
